% ============================================================
%  Bibliographie — Analyse des Données
% ============================================================
\section*{Bibliographie}

\begin{frame}[allowframebreaks]{Références}

\textbf{\large Ouvrages fondamentaux (école française)}\\[6pt]

\begin{enumerate}
  \item \textbf{Saporta, G.} (2011).
        \textit{Probabilités, analyse des données et statistique} (3e éd.).
        Paris : Technip. \emph{(Référence francophone de référence.)}

  \item \textbf{Lebart, L., Morineau, A., \& Piron, M.} (2006).
        \textit{Statistique exploratoire multidimensionnelle} (4e éd.).
        Paris : Dunod.

  \item \textbf{Escofier, B., \& Pagès, J.} (2008).
        \textit{Analyses factorielles simples et multiples : objectifs,
        méthodes et interprétation} (4e éd.). Paris : Dunod.

  \item \textbf{Husson, F., Lê, S., \& Pagès, J.} (2017).
        \textit{Analyse de données avec R} (2e éd.).
        Rennes : Presses Universitaires de Rennes.

  \item \textbf{Cornillon, P.-A., Guyader, A., Husson, F. et al.} (2018).
        \textit{Statistique avec R} (4e éd.). Rennes : PUR.
\end{enumerate}

\framebreak

\textbf{\large Auteurs historiques}\\[6pt]

\begin{enumerate}
  \setcounter{enumi}{5}
  \item \textbf{Benzécri, J.-P.} (1973).
        \textit{L'analyse des données} (2 volumes : I — Taxinomie ;
        II — L'analyse des correspondances). Paris : Dunod.
        \emph{(Texte fondateur de l'école française.)}

  \item \textbf{Cazes, P.} (1990).
        Méthodes de \textit{Statistique Descriptive Multidimensionnelle}.
        \textit{Cahiers du Bureau Universitaire de Recherche Opérationnelle}.

  \item \textbf{Cailliez, F., \& Pagès, J.-P.} (1976).
        \textit{Introduction à l'analyse des données}.
        Paris : SMASH.

  \item \textbf{Lebart, L., \& Fenelon, J.-P.} (1971).
        \textit{Statistique et informatique appliquées}.
        Paris : Dunod.

  \item \textbf{Bouroche, J.-M., \& Saporta, G.} (1980).
        \textit{L'analyse des données} (\textit{Que sais-je ?}).
        Paris : PUF.
\end{enumerate}

\framebreak

\textbf{\large Manuels internationaux}\\[6pt]

\begin{enumerate}
  \setcounter{enumi}{10}
  \item \textbf{Hastie, T., Tibshirani, R., \& Friedman, J.} (2009).
        \textit{The Elements of Statistical Learning} (2e éd.).
        New York : Springer.

  \item \textbf{Mardia, K. V., Kent, J. T., \& Bibby, J. M.} (1979).
        \textit{Multivariate Analysis}. London : Academic Press.

  \item \textbf{Jolliffe, I. T.} (2002).
        \textit{Principal Component Analysis} (2e éd.).
        New York : Springer.

  \item \textbf{Greenacre, M.} (2017).
        \textit{Correspondence Analysis in Practice} (3e éd.).
        Boca Raton : CRC Press.

  \item \textbf{Izenman, A. J.} (2008).
        \textit{Modern Multivariate Statistical Techniques}.
        New York : Springer.
\end{enumerate}

\framebreak

\textbf{\large Articles séminaux — ACP et SVD}\\[6pt]

\begin{enumerate}
  \setcounter{enumi}{15}
  \item \textbf{Pearson, K.} (1901).
        On lines and planes of closest fit to systems of points in space.
        \textit{Philosophical Magazine}, 2(11), 559--572. \emph{(Origine de l'ACP.)}

  \item \textbf{Hotelling, H.} (1933).
        Analysis of a complex of statistical variables into principal components.
        \textit{Journal of Educational Psychology}, 24, 417--441 et 498--520.

  \item \textbf{Eckart, C., \& Young, G.} (1936).
        The approximation of one matrix by another of lower rank.
        \textit{Psychometrika}, 1(3), 211--218.

  \item \textbf{Mirsky, L.} (1960).
        Symmetric gauge functions and unitarily invariant norms.
        \textit{Quarterly Journal of Mathematics}, 11(1), 50--59.

  \item \textbf{Jolliffe, I. T., \& Cadima, J.} (2016).
        Principal component analysis: a review and recent developments.
        \textit{Philosophical Transactions of the Royal Society A}, 374.
\end{enumerate}

\framebreak

\textbf{\large Articles séminaux — AFC, ACM, classification}\\[6pt]

\begin{enumerate}
  \setcounter{enumi}{20}
  \item \textbf{Benzécri, J.-P.} (1969).
        Statistical analysis as a tool to make patterns emerge from data.
        \textit{Methodologies of Pattern Recognition}, 35--74.

  \item \textbf{Greenacre, M., \& Blasius, J.} (Eds.) (2006).
        \textit{Multiple Correspondence Analysis and Related Methods}.
        Boca Raton : Chapman \& Hall/CRC.

  \item \textbf{Ward, J. H.} (1963).
        Hierarchical grouping to optimize an objective function.
        \textit{Journal of the American Statistical Association}, 58(301), 236--244.

  \item \textbf{MacQueen, J.} (1967).
        Some methods for classification and analysis of multivariate observations.
        \textit{Berkeley Symposium on Mathematical Statistics}, 1, 281--297.

  \item \textbf{Arthur, D., \& Vassilvitskii, S.} (2007).
        k-means++: The advantages of careful seeding.
        \textit{Proc. SODA 2007}, 1027--1035.

  \item \textbf{Rousseeuw, P. J.} (1987).
        Silhouettes: A graphical aid to the interpretation and validation of
        cluster analysis. \textit{Journal of Computational and Applied Mathematics},
        20, 53--65.
\end{enumerate}

\framebreak

\textbf{\large Articles séminaux — AFD et méthodes avancées}\\[6pt]

\begin{enumerate}
  \setcounter{enumi}{26}
  \item \textbf{Fisher, R. A.} (1936).
        The use of multiple measurements in taxonomic problems.
        \textit{Annals of Eugenics}, 7(2), 179--188. \emph{(Origine de la LDA.)}

  \item \textbf{Hotelling, H.} (1936).
        Relations between two sets of variates.
        \textit{Biometrika}, 28(3/4), 321--377. \emph{(Analyse canonique.)}

  \item \textbf{Torgerson, W. S.} (1952).
        Multidimensional scaling: I. Theory and method.
        \textit{Psychometrika}, 17, 401--419.

  \item \textbf{Kruskal, J. B.} (1964).
        Multidimensional scaling by optimizing goodness of fit to a nonmetric hypothesis.
        \textit{Psychometrika}, 29(1), 1--27.

  \item \textbf{Wold, H.} (1975).
        Soft modelling by latent variables: The non-linear iterative partial
        least squares (NIPALS) approach. \textit{Perspectives in Probability
        and Statistics}, 117--142. \emph{(Régression PLS.)}

  \item \textbf{Schölkopf, B., Smola, A., \& Müller, K.-R.} (1998).
        Nonlinear component analysis as a kernel eigenvalue problem.
        \textit{Neural Computation}, 10(5), 1299--1319.

  \item \textbf{Tenenbaum, J. B., de Silva, V., \& Langford, J. C.} (2000).
        A global geometric framework for nonlinear dimensionality reduction.
        \textit{Science}, 290(5500), 2319--2323. \emph{(Isomap.)}

  \item \textbf{Roweis, S. T., \& Saul, L. K.} (2000).
        Nonlinear dimensionality reduction by locally linear embedding.
        \textit{Science}, 290(5500), 2323--2326. \emph{(LLE.)}

  \item \textbf{Candès, E. J., Li, X., Ma, Y., \& Wright, J.} (2011).
        Robust principal component analysis ?
        \textit{Journal of the ACM}, 58(3), 1--37.
\end{enumerate}

\framebreak

\textbf{\large Méthodes contemporaines}\\[6pt]

\begin{enumerate}
  \setcounter{enumi}{35}
  \item \textbf{Zou, H., Hastie, T., \& Tibshirani, R.} (2006).
        Sparse principal component analysis.
        \textit{Journal of Computational and Graphical Statistics}, 15(2), 265--286.

  \item \textbf{van der Maaten, L., \& Hinton, G.} (2008).
        Visualizing data using t-SNE.
        \textit{Journal of Machine Learning Research}, 9, 2579--2605.

  \item \textbf{McInnes, L., Healy, J., \& Melville, J.} (2018).
        UMAP: Uniform manifold approximation and projection for dimension
        reduction. \textit{arXiv:1802.03426}.

  \item \textbf{Ramsay, J. O., \& Silverman, B. W.} (2005).
        \textit{Functional Data Analysis} (2e éd.). New York : Springer.

  \item \textbf{Kolda, T. G., \& Bader, B. W.} (2009).
        Tensor decompositions and applications.
        \textit{SIAM Review}, 51(3), 455--500.

  \item \textbf{Carlsson, G.} (2009).
        Topology and data. \textit{Bulletin of the AMS}, 46(2), 255--308.
        \emph{(TDA.)}
\end{enumerate}

\framebreak

\textbf{\large Logiciels et packages R}\\[6pt]

\begin{enumerate}
  \setcounter{enumi}{41}
  \item \textbf{Lê, S., Josse, J., \& Husson, F.} (2008).
        FactoMineR: An R package for multivariate analysis.
        \textit{Journal of Statistical Software}, 25(1), 1--18.
        \texttt{factominer.free.fr}

  \item \textbf{Kassambara, A., \& Mundt, F.} (2020).
        \textit{factoextra: Extract and Visualize the Results of Multivariate
        Data Analyses}. R package version 1.0.7.

  \item \textbf{Dray, S., \& Dufour, A.-B.} (2007).
        The ade4 package: implementing the duality diagram for ecologists.
        \textit{Journal of Statistical Software}, 22(4), 1--20.

  \item \textbf{Venables, W. N., \& Ripley, B. D.} (2002).
        \textit{Modern Applied Statistics with S} (4e éd.). Springer.
        \emph{(Package \texttt{MASS} : lda, qda, isoMDS.)}

  \item \textbf{R Core Team} (2024).
        \textit{R: A language and environment for statistical computing}.
        R Foundation for Statistical Computing, Vienna.
        \texttt{www.R-project.org}
\end{enumerate}

\framebreak

\textbf{\large Ressources en ligne et communauté}

\vspace{0.3cm}

\begin{block}{Sites web et MOOC}
\footnotesize
\begin{itemize}
  \item \textbf{factominer.free.fr} — site officiel FactoMineR (Husson \& Lê)
  \item \textbf{rdrr.io/cran/FactoMineR} — documentation interactive
  \item \textbf{r-graph-gallery.com} — visualisations factorielles
  \item \textbf{MOOC \emph{Analyse des Données Multidimensionnelles}}
        (FUN, Husson \& Lê, AGROCAMPUS Ouest)
  \item \textbf{distill.pub} — articles visuels sur t-SNE, UMAP
\end{itemize}
\end{block}

\vspace{0.3cm}

\begin{block}{Conférences clés}
\footnotesize
\begin{itemize}
  \item \textbf{CARME} — Conferences on Correspondence Analysis and Related Methods
  \item \textbf{IFCS} — International Federation of Classification Societies
  \item \textbf{Rencontres R} (France) — communauté R francophone
\end{itemize}
\end{block}

\end{frame}
